Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-07 index.qmd
section:background,flow; process:write,figures

<!-- 4. But which approach are we going to use? -->
However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020]. DAGs serve as heuristics, effectively conveying the assumed causal structure of a system. A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].

@@

<!-- 2. What does the structural approach uses?, and what are DAGs? -->
In statistics, one application incorporating concepts from graph theory is causal inference. Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) to provide a graphical and formal representation of the causal structure of a system [@Neal_2020]. In this context, a *graph* denotes a collection of nodes connected by edges, where nodes represent random variables. The term *directed* indicates the edges of the graph extend from one node to another, with arrows showing the direction of causal influence. Moreover, the term *acyclic* indicates the causal influences do not form a loop, meaning the influences do not cycle back on themselves [@McElreath_2020].

<!-- 3. What are the advantages of DAGs? -->
DAGs offer two key advantages for modeling causality. Firstly, DAGs represent causal relations in a nonparametric and fully interactive manner. This feature implies that feasible causal analysis strategies usually do not require specifying the type of data or the nature of the functional dependence among variables [@Morgan_et_al_2014]. Secondly, regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This further implies that to understand the causal assumptions of a complex system, one can decompose a structure into these basic building blocks and analyze the causal assumptions associated with each block individually [@McElreath_2020]. Importantly, these building blocks can be represented in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].

<!-- 3.2 the simplified DAG -->
Often, the DAGs omit the exogenous variables for simplicity, resulting in the *standard* representation. However, including exogenous variables in a graph can be beneficial in some scenarios, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts explored in the following section. The standard representation is illustrated in the middle panels of @fig-dags_scms.

@@

<!-- 3.3 the SCM -->
Lastly, the right panels of @fig-dags_scms depict the SCM form of the five fundamental building blocks. SCMs are formal mathematical models that represent causal relationships within a system or population [@Hitchcock_2024]. An SCM is defined by a set of endogenous variables $V$, a set of exogenous variables $E$, and a set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Neal_2020]. These functions, referred to as structural equations, specify each endogenous variable as nonparametric functions of other variables. Moreover, SCMs use the symbol '$:=$' to indicate the variables' asymmetrical causal dependence and the symbol '$\dsep$' to represent *d-separation*, which roughly equates to the concept of variable independence. The concepts of d-separation and causal (in)dependence are explored in the following section.

@@

<!-- 4. But what does the building blocks mean? -->
A careful examination of @fig-dags_scms highlights the assumptions underlying these building blocks. Figures [-@fig-mdag_bb1], [-@fig-sdag_bb1], and SCM [-@fig-scm_bb1] depict two unconnected nodes, representing a scenario where variables $X$ and $Y$ are not causally related. Figures [-@fig-mdag_bb2], [-@fig-sdag_bb2], and SCM [-@fig-scm_bb2] illustrate two connected nodes, showing a scenario where a *parent* node $X$ exerts a causal influence on a *child* node $Y$. Consequently, $Y$ is considered a *descendant* of $X$. Figures [-@fig-mdag_bb3], [-@fig-sdag_bb3], and SCM [-@fig-scm_bb3] depict a *chain* or *pipe*, where $X$ influences $Z$, and $Z$ influences $Y$. In this configuration, $X$ is a parent node of $Z$, and $Z$ is a parent node of $Y$. Furthermore, the DAGs show that $X$ is an *ancestor* of $Y$ and that $Z$ fully *mediates* the relationship between these variables. Figures [-@fig-mdag_bb4], [-@fig-sdag_bb4], and SCM [-@fig-scm_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$. Here, $Z$ is a parent node of $X$ and $Y$. Finally, Figures [-@fig-mdag_bb5], [-@fig-sdag_bb5], SCM [-@fig-scm_bb5] depict a *collider*, also known as *inmorality*, where variables $X$ and $Y$ are concurrent causes of $Z$. In this configuration, $X$ and $Y$ are not causally related to each other but both influence $Z$. Additionally, in all SCMs, the errors are assumed to be mutually independent of each other and of all other variables in the graph, as evidenced by the pairwise relations $e_{X} \dsep e_{Y}$, $e_{X} \dsep e_{Z}$, and $e_{Z} \dsep e_{Y}$.

<!-- 5. A motivating example -->
Given the heuristic nature of DAGs, a motivating example can help clarify how to use the five fundamental building blocks to construct the causal structure of a system. Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs investigation. The problem suggests that a variable $X$ potentially influences both $T$ and $Y$. Beyond these relationships, the problem does not specify any further variables of interest. Such scenarios are not hard to imagine. For instance, $T$ might represent different treatments that could affect the recovery from cancer $Y$, while $X$ could denote the patient's age. Similarly, in the context of a DCJ study like the one described by @Boonen_et_al_2020, $T$ could represent the duration of a child's cochlear implant use, which might influence the child's overall speech quality $Y$, with $X$ indicating the child's hearing status. [(not a bad example, but I prefer one using writing skills)]{style="color:blue;"}

@@

<!-- 1. But what are the assumption that allow us to treat DAGs as causal structures -->
![ The flow of association and causation in graphs. Extracted from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=100%}

@@

@@

## But where does it all fit? {#sec-background-where}

@@

<!-- Initially, it defines the different targets of inference (i.e., the *estimands*), which form the basis for evaluating the reliability of the DCJ method. The targets  Subsequently, the research constructs a *scientific model* for the DCJ procedure, elucidating the causal assumptions and mechanisms underlying the system, aiding in drawing inferences about causal relationships from DCJ data. Next, the study translates the scientific model into a probabilistic statistical model (i.e., an *estimator*) to derive a *statistical estimand* for the various inference targets. Bayesian statistical methods offer an intuitive approach for the task. Finally, employing multiple simulated datasets for different combinations of the relevant assessment factors, the study generates multiple *estimates* or approximations for the statistical estimands. Notably, the statistical estimands in simulated data are known, enabling the study to assess the reliability of DCJ by evaluating the *proximity* of these estimands to the generated estimates, resembling a power analysis. -->

@@

>>>>

<!-- 4. But which approach are we going to use? -->
However, the structural approach offers an additional key advantage over the potential outcomes approach: it enables the graphical representation of any system through directed acyclic graphs [DAG, @Gross_et_al_2018; @Neal_2020]. DAGs serve as heuristics, effectively conveying the assumed causal structure of a system [@Pearl_et_al_2016]. A heuristic is a strategy that simplifies information to make decisions more quickly, efficiently, and sometimes more accurately than complex methods [@Chow_2015]. Consequently, DAGs do not represent detailed statistical models but allow researchers to deduce which statistical models can provide valid causal inferences, assuming the causal structure depicted in the DAGs are accurate [@McElreath_2020].

@@

<!-- 2. What does the structural approach uses?, and what are DAGs? -->
In statistics, one application incorporating concepts from graph theory is causal inference. Specifically, the structural approach to causal inference uses directed acyclic graphs (DAG) to provide a graphical and formal representation of the causal structure of a system [@Neal_2020]. In this context, a *graph* denotes a collection of nodes connected by edges, where nodes represent random variables. The term *directed* indicates the edges of the graph extend from one node to another, with arrows showing the direction of causal influence. Moreover, the term *acyclic* indicates the causal influences do not form a loop, meaning the influences do not cycle back on themselves [@McElreath_2020]. Several examples of DAGs can be found in @fig-dags_scms.

<!-- 3. What are the advantages of DAGs? -->
DAGs offer two key advantages for modeling causal structures. Firstly, they represent causal relations in a nonparametric and fully interactive manner. This feature allows for feasible causal analysis strategies without needing the specification of the type of data or the nature of the functional dependence among variables [@Morgan_et_al_2014]. Secondly, regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature enables the decomposition of complex structures into basic building blocks, facilitating the analysis of these structures by focusing on the causal assumptions associated with each building block individually [@McElreath_2020]. These building blocks can be represented in three ways: the magnified representation, the standard representation, and the structural causal model form [SCM, @Morgan_et_al_2014].

@@

@@

<!-- 3.2 the simplified DAG -->
Often, DAGs omit the exogenous variables for simplicity, resulting in the *standard* representation. However, including exogenous variables in a graph can be beneficial in some scenarios, as their presence can reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020], concepts explored in the following section. The standard representation is illustrated in the middle panels of @fig-dags_scms.

<!-- 3.3 the SCM -->
Lastly, the right panels of @fig-dags_scms depict the SCM form of the fundamental building blocks. SCMs are formal mathematical models defined by a set of endogenous variables $V$, a set of exogenous variables $E$, and a set of functions $F=\{f_{X},f_{Z},f_{Y}\}$ [@Pearl_2009; @Neal_2020]. These functions, referred to as structural equations, specify each endogenous variable as nonparametric functions of other variables. Moreover, SCMs use the symbol '$:=$' to indicate the variables' asymmetrical causal dependence and the symbol '$\dsep$' to represent *d-separation*, which roughly equates to the concept of variable independence. The concepts of d-separation and causal (in)dependence are explored in the following section.

<!-- 4. But what does the building blocks mean? -->
A careful examination of @fig-dags_scms highlights the assumptions underlying these building blocks. Figures [-@fig-mdag_bb1], [-@fig-sdag_bb1], and SCM [-@fig-scm_bb1] depict two unconnected nodes, representing a scenario where variables $X$ and $Y$ are not causally related. Figures [-@fig-mdag_bb2], [-@fig-sdag_bb2], and SCM [-@fig-scm_bb2] illustrate two connected nodes, showing a scenario where a *parent* node $X$ exerts a causal influence on a *child* node $Y$. Consequently, $Y$ is considered a *descendant* of $X$. Figures [-@fig-mdag_bb3], [-@fig-sdag_bb3], and SCM [-@fig-scm_bb3] depict a *chain* or *pipe*, where $X$ influences $Z$, and $Z$ influences $Y$. In this configuration, $X$ is a parent node of $Z$, and $Z$ is a parent node of $Y$. This creates a *directed path* between $X$ and $Y$. Consequently, $X$ is an *ancestor* of $Y$, and $Z$ fully *mediates* the relationship between the two. Figures [-@fig-mdag_bb4], [-@fig-sdag_bb4], and SCM [-@fig-scm_bb4] illustrate a *fork*, where variables $X$ and $Y$ are both influenced by $Z$. Here, $Z$ is a parent node of $X$ and $Y$. Finally, Figures [-@fig-mdag_bb5], [-@fig-sdag_bb5], SCM [-@fig-scm_bb5] depict a *collider*, also known as *inmorality*, where variables $X$ and $Y$ are concurrent causes of $Z$. In this configuration, $X$ and $Y$ are not causally related to each other but both influence $Z$. Additionally, in all SCMs, the errors are assumed to be mutually independent of each other and of all other variables in the graph, as evidenced by the pairwise relations $e_{X} \dsep e_{Y}$, $e_{X} \dsep e_{Z}$, and $e_{Z} \dsep e_{Y}$.

@@

<!-- 1. What has been said about DAGs and SCMs -->
<!-- So far, the explanation of DAGs and SCMs has been fairly concise, sufficient to emphasize their effectiveness in encoding causal structures. However, nothing has been said about the assumptions required to use DAGs and SCMs as encoders of causal assumptions. This is important because, while encoding causal structures with DAGs can be an intriguing exercise in itself, the primary goal of an investigator is to address a research problem. -->

<!-- Specifically, the investigator must define tthe different targets of inference (i.e., the *estimands*), which form the basis for evaluating the reliability of the DCJ method. The targets  Subsequently, the research constructs a *scientific model* for the DCJ procedure, elucidating the causal assumptions and mechanisms underlying the system, aiding in drawing inferences about causal relationships from DCJ data. Next, the study translates the scientific model into a probabilistic statistical model (i.e., an *estimator*) to derive a *statistical estimand* for the various inference targets. Bayesian statistical methods offer an intuitive approach for the task. Finally, employing multiple simulated datasets for different combinations of the relevant assessment factors, the study generates multiple *estimates* or approximations for the statistical estimands. Notably, the statistical estimands in simulated data are known, enabling the study to assess the reliability of DCJ by evaluating the *proximity* of these estimands to the generated estimates, resembling a power analysis. -->

<!-- translate the encoded causal information into a probabilistic statistical model. Then, using this model along with data, the investigator must produce an estimate that addresses the research problem of interest. This procedure is illustrated in the Identification-Estimation flowchart presented in @fig-IEflow. -->

@@

<!-- 1. But what are the assumption that allow us to treat DAGs as causal structures -->
![ The flow of association and causation in graphs. Extracted from @Neal_2020 [pp. 31] ](images/figures/ACflow.png){#fig-ACflow fig-align="center" width=90%}

<!-- Another advantage of graphical models is that they allow us to express joint distributions very efficiently. -->

<!-- Second, in a fully specified SCM, we can represent the joint distributions of n variables -->
<!-- with greater efficiency: We need only to specify the n functions that govern the relationships between the variables, and then from the probabilities of the error terms, we can discover all the probabilities that govern the joint distribution. -->

@@

@@

@@

## A motivating example {#sec-framework-example}

Given the heuristic nature of DAGs, a motivating example can help clarify how to use the five fundamental building blocks to construct the causal structure of a system. Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs investigation. The problem suggests that a variable $X$ potentially influences both $T$ and $Y$. Beyond these relationships, the problem does not specify any further variables of interest. Such scenarios are not hard to imagine. For instance, $T$ might represent different treatments that could affect the recovery from cancer $Y$, while $X$ could denote the patient's age. Similarly, in the context of a DCJ study like the one described by @Boonen_et_al_2020, $T$ could represent the duration of a child's cochlear implant use, which might influence the child's overall speech quality $Y$, with $X$ indicating the child's hearing status. [(not a bad example, but I prefer one using writing skills)]{style="color:blue;"}
